---
title: "Capital BikeShare Analysis"
author: "Dataphilers"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
```


Read the CSV, for only last 5 years i.e. September 2013 to September 2019 (Note there should be 3000 data points)
```{r read_csv, message=FALSE, warning=FALSE}
# due to storage limit of bitbucket, download Capital Bike Data file from https://www.capitalbikeshare.com/system-data
bikeFileNames <- list.files(path="Capital Bike Share Data\\2019",pattern="*.csv",full.names=T) #change this [2013,2014,2015,2016,2017,2018,2019]
tbl <- sapply(bikeFileNames, read.csv, simplify=FALSE) %>% bind_rows()
```

* Convert into appropriate data type
  * Duration rename column to seconds
```{r}
colnames(tbl)[colnames(tbl)=='Duration'] = 'seconds'
```
  
  * start.date and end.date convert to POSIXct type
```{r}
# x <- strptime("2019-10-29 00:04:23",format="%Y-%m-%d %H:%M:%S",tz = "EST")
# print(x$mday)
# print(x$mon+1)
# print(x$year+1900)
# print(x$hour)
# print(x$min)
# print(x$sec)
# print(x$wday) # 0=Sunday, 1=Monday ....
# print(x$zone)

tbl$Start.date <- strptime(tbl$Start.date,format="%Y-%m-%d %H:%M:%S",tz = "EST")
tbl$End.date <- strptime(tbl$End.date,format="%Y-%m-%d %H:%M:%S",tz = "EST")
```

  * Start.station.number convert to factor
  * Start.station convert to factor
  * End.station.number convert to factor
  * End.station  convert to factor
  * Bike.number convert to factor
  * member.type convert to factor




* add weather data
```{r}
#read all the columns as chr class
weatherData <- read.csv("Capital Bike Share Data/WeatherData/aggregatedWeatherData.csv",header = T,stringsAsFactors = FALSE)
```
  
  * combine weatherData with tbl to get weather information for start Date and End date columns based on common column
```{r}
#add temp column in tbl
tbl$temp <- format(round(tbl$Start.date, units="hours"), format="%Y-%m-%d %H:%M:%S")
```

```{r}
#merge the 2 dataframes
combined <- merge(tbl,weatherData,by.x = "temp",by.y = "Time")
```

* Weekend or Weekday
  * using existing tbl column add new column which will contain whether date is weekday or weekend
```{r}
combined$weekday <- ifelse(combined$Start.date$wday==0 | combined$Start.date$wday==6, "Weekend",  "Weekday")
```

* add holiday data from holiday list csv
```{r}
holidayData <- read.csv("Capital Bike Share Data/Holiday List/usholidays.csv",header = T,stringsAsFactors = FALSE)
holidayData <- holidayData[c("Date","Holiday")]

# reformat the temp column of combined dataframe to be able to merge with holidayData dataframe
combined$temp <- format(combined$Start.date, format="%Y-%m-%d")

#merge holidays
combined <- merge(combined,holidayData,by.x = "temp",by.y = "Date",all.x = TRUE)
```

* Time of day => working or non working hour 
```{r}
# https://en.wikipedia.org/wiki/9_to_5_(disambiguation)
combined$timeOfDay <- ifelse(combined$Start.date$hour>=9 & combined$Start.date$hour<=17,
                             "Working Hour",
                             "Non Working Hour"
                             )
```


* Season => Fall, winter, summer, spring 
```{r}
# https://www.visittheusa.com/info/weather-seasons
# Summer: June - August
# Fall: September - November
# Winter: December - February
# Spring: March - May 
# months go from 0 to 11
combined$Start.date <- strptime(combined$Start.date,format="%Y-%m-%d %H:%M:%S",tz = "EST")

combined$season <- ifelse(combined$Start.date$mon %in% c(11,0,1), 
                            "Winter",
                            ifelse(combined$Start.date$mon %in% c(2,3,4), 
                                   "Spring",
                                   ifelse(combined$Start.date$mon %in% c(5,6,7), 
                                          "Summer",
                                          "Fall")
                                   )
                          )
```




```{r}
n <- c("Start.date"          
       ,"Start.station"       
       ,"Condition"            
       ,"Wind"                
       ,"Temperature"          
       ,"Dew.Point"            
       ,"Humidity"            
       ,"Wind.Speed"           
       ,"Wind.Gust"            
       ,"Pressure"            
       ,"Precip."              
       ,"Holiday"              
       ,"weekday"             
       ,"timeOfDay"            
       ,"season")
combined <- combined[n]
```


```{r}
colnames(combined)[colnames(combined)=="Temperature"]="Temperature (F)"
colnames(combined)[colnames(combined)=="Dew.Point"]="Dew.Point (F)"
colnames(combined)[colnames(combined)=="Humidity"]="Humidity (%)"
colnames(combined)[colnames(combined)=="Wind.Speed"]="Wind.Speed (mph)"
colnames(combined)[colnames(combined)=="Wind.Gust"]="Wind.Gust (mph)"
colnames(combined)[colnames(combined)=="Pressure"]="Pressure (in)"
colnames(combined)[colnames(combined)=="Precip."]="Precip. (in)"
```

```{r}
combined$Start.date <- format(round(combined$Start.date, units="hours"), format="%Y-%m-%d %H:%M:%S")
```

Group by Start Date and Start.station
```{r}
combined <- combined %>% group_by(Start.date,Start.station) %>% mutate(noOfBikes = n()) %>% distinct(Start.date,Start.station,.keep_all = TRUE) %>% arrange(Start.station)
```

Write to csv, finally!
```{r}
write.csv(combined,"rawCapitalBikeShare2019.csv") #change this [2013,2014,2015,2016,2017,2018,2019]
```


Combining all the csv back
```{r}
calendarNames <- list.files(path=".",pattern="*.csv",full.names=T)
finalDF <- sapply(calendarNames, read.csv, simplify=FALSE) %>% bind_rows()
write.csv(finalDF,"rawCapitalBikeShare.csv")
```

